The Changing Face of Pediatric Pulmonary Exacerbations in CF (NACFC 2023)

### 2023-10-12 Scott Sagel/PEx in CF/nacfc_2023.qmd
rough draft of nacfc presentation done

---
title: "The Changing Face of Pediatric Pulmonary Exacerbations in CF (NACFC 2023)"
author: "Tim Vigers"
date: "today"
format:
  html:
    toc: false
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
editor: source
execute: 
  echo: false
---

```{r}
#| include: false
library(tidyverse)
library(gtsummary)
# source("~/GitHub/CF/Scott Sagel/PEx in CF/create_pex_analysis_dataset.R")
load("~/Dropbox/Work/Vigers/CF/Scott Sagel/PEx in CF/Data_Cleaned/analysis_dataset.RData")
```

@@

The number of hospitalizations for PEx decreased from `r sum(df$Year==2018)` in 2018 to `r sum(df$Year==2022)` in 2022.

# Summary Statistics

```{r}
#| label: tbl-all
#| tbl-cap: Summary by Year
#| message: false
summary_table %>%
  select(-sid, -redcap_data_access_group) %>%
  select(num_hosp, days_admit:ppfev1_diff, age_at_admit, everything()) %>%
  tbl_summary(
    by = admit_year,
    type = list(num_hosp ~ "continuous"),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      num_hosp ~ "{median} [{p25}, {p75}]",
      days_admit ~ "{median} [{p25}, {p75}]"
    ),
    missing_text = "(Missing)",
    label = labels
  )
```

```{r}
#| label: tbl-site
#| tbl-cap: Stratified by Site
#| message: false
summary_table %>%
  select(-sid, -cftr_mod_spec) %>%
  select(num_hosp, days_admit:ppfev1_diff, age_at_admit, everything()) %>%
  tbl_strata(
    strata = redcap_data_access_group,
    .tbl_fun =
      ~ .x %>%
        tbl_summary(
          by = admit_year,
          type = list(num_hosp ~ "continuous"),
          statistic = list(
            all_continuous() ~ "{mean} ({sd})",
            num_hosp ~ "{median} [{p25}, {p75}]",
            days_admit ~ "{median} [{p25}, {p75}]"
          ),
          missing_text = "(Missing)",
          label = labels
        )
  )
```

>>>>

---
title: "The Changing Face of Pediatric Pulmonary Exacerbations in CF (NACFC 2023)"
author: "Tim Vigers"
date: "today"
format:
  html:
    toc: false
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
editor: source
---

```{r}
#| include: false
library(tidyverse)
library(gtsummary)
# source("~/GitHub/CF/Scott Sagel/PEx in CF/create_pex_analysis_dataset.R")
load("~/Dropbox/Work/Vigers/CF/Scott Sagel/PEx in CF/Data_Cleaned/analysis_dataset.RData")
# Colors for plots
yellow <- "#f8b80a"
red <- "#e30022"
purple <- "#7b2670"
blue <- "#022451"
```

@@

# Summary Statistics (Per Person)

```{r}
#| label: tbl-all
#| tbl-cap: Summary by Year
#| message: false
summary_table %>%
  select(-sid, -redcap_data_access_group) %>%
  select(num_hosp, days_admit:ppfev1_diff, age_at_admit, everything()) %>%
  tbl_summary(
    by = admit_year,
    type = list(num_hosp ~ "continuous"),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      num_hosp ~ "{median} [{p25}, {p75}]",
      days_admit ~ "{median} [{p25}, {p75}]"
    ),
    missing_text = "(Missing)",
    label = labels
  ) %>%
  add_p(
    test = list(
      all_continuous() ~ "t.test",
      num_hosp ~ "wilcox.test",
      days_admit ~ "wilcox.test"
    ),
    test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)
  ) %>%
  separate_p_footnotes()
```

```{r}
#| label: tbl-site
#| tbl-cap: Stratified by Site
#| message: false
summary_table %>%
  select(-sid, -cftr_mod_spec) %>%
  select(num_hosp, days_admit:ppfev1_diff, age_at_admit, everything()) %>%
  tbl_strata(
    strata = redcap_data_access_group,
    .tbl_fun =
      ~ .x %>%
        tbl_summary(
          by = admit_year,
          type = list(num_hosp ~ "continuous"),
          statistic = list(
            all_continuous() ~ "{mean} ({sd})",
            num_hosp ~ "{median} [{p25}, {p75}]",
            days_admit ~ "{median} [{p25}, {p75}]"
          ),
          missing_text = "(Missing)",
          label = labels
        ) %>%
        add_p(
          test = list(
            all_continuous() ~ "t.test",
            num_hosp ~ "wilcox.test",
            days_admit ~ "wilcox.test"
          ),
          test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)
        ) %>%
        separate_p_footnotes()
  )
```

# Plots

## Slide 9

```{r}
# Number of hospitalizations
df %>%
  group_by(redcap_data_access_group, admit_year) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = redcap_data_access_group, y = n, fill = factor(admit_year))) +
  geom_bar(position = "dodge", stat = "identity") +
  xlab("Site") +
  ylab("Num. of Hospitalizations") +
  theme_bw() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = c(purple, yellow))
# % hospitalized 2 or more times in both years
df %>%
  group_by(sid, redcap_data_access_group, admit_year) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(redcap_data_access_group, admit_year) %>%
  summarise(perc = mean(n >= 2) * 100, .groups = "drop") %>%
  ggplot(aes(x = redcap_data_access_group, y = perc, fill = factor(admit_year))) +
  geom_bar(position = "dodge", stat = "identity") +
  xlab("Site") +
  ylab("Percent With >= 2 Hospitalizations") +
  theme_bw() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = c(purple, yellow))
```

## Slide 10

For the box and whisker plot of age at admission, we used the average age for people who were admitted more than once in a given year.

```{r}
#| message: false
#| warning: false
# Age
df %>%
  group_by(sid, redcap_data_access_group, admit_year) %>%
  summarise(age_at_admit = mean(age_at_admit, na.rm = T), .groups = "drop") %>%
  ggplot(aes(x = redcap_data_access_group, y = age_at_admit, 
             fill = factor(admit_year))) +
  geom_boxplot() +
  xlab("Site") +
  ylab("Age at Admission") +
  theme_bw() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = c(purple, red))
# Sex
df %>%
  group_by(redcap_data_access_group, admit_year, sid) %>%
  summarise(sex = unique(sex)) %>%
  summarise(Male = mean(sex == "Male") * 100, Female = mean(sex == "Female") * 100) %>%
  pivot_longer(c(Male, Female), names_to = "sex", values_to = "perc") %>%
  ggplot(aes(x = factor(admit_year), y = perc, fill = factor(sex))) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(aes(label = paste(round(perc, 1), "%")),
    position = position_stack(vjust = 0.5), size = 2
  ) +
  ylab("%") +
  facet_grid(~redcap_data_access_group) +
  theme_bw() +
  theme(legend.title = element_blank(), axis.title.x = element_blank()) +
  scale_fill_manual(values = c(yellow, red))
# Race
df %>%
  group_by(redcap_data_access_group, admit_year, sid) %>%
  summarise(race = unique(race)) %>%
  summarise(
    White = mean(race == "White") * 100,
    `Black or African-American` = mean(race == "Black or African-American") * 100,
    Asian = mean(race == "NE Asian") * 100,
    `Other/mixed` = mean(race == "Other/mixed") * 100
  ) %>%
  pivot_longer(c(
    White, `Black or African-American`,
    Asian, `Other/mixed`
  ), names_to = "race", values_to = "perc") %>%
  ggplot(aes(x = factor(admit_year), y = perc, fill = factor(race))) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(aes(label = paste(round(perc, 1), "%")),
    position = position_stack(vjust = 0.5), size = 2
  ) +
  ylab("%") +
  facet_grid(~redcap_data_access_group) +
  theme_bw() +
  theme(legend.title = element_blank(), axis.title.x = element_blank()) +
  scale_fill_manual(values = c(red, purple, blue, yellow))
# Ethnicity
df %>%
  group_by(redcap_data_access_group, admit_year, sid) %>%
  summarise(ethnicity = unique(ethnicity)) %>%
  summarise(
    `Hispanic or Latino` = mean(ethnicity == "Hispanic or Latino") * 100,
    `Not Hispanic or Latino` = mean(ethnicity == "Not Hispanic or Latino") * 100,
    `Unknown or not reported` = mean(ethnicity == "Unknown or not reported") * 100
  ) %>%
  pivot_longer(c(
    `Hispanic or Latino`, `Not Hispanic or Latino`, `Unknown or not reported`
  ), names_to = "ethnicity", values_to = "perc") %>%
  ggplot(aes(x = factor(admit_year), y = perc, fill = factor(ethnicity))) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(aes(label = paste(round(perc, 1), "%")),
    position = position_stack(vjust = 0.5), size = 2
  ) +
  ylab("%") +
  facet_grid(~redcap_data_access_group) +
  theme_bw() +
  theme(legend.title = element_blank(), axis.title.x = element_blank()) +
  scale_fill_manual(values = c(red, purple, yellow))
```

## Slide 11

```{r}
#| message: false
# Ethnicity
df %>%
  group_by(redcap_data_access_group, admit_year, sid) %>%
  summarise(genotype = unique(genotype)) %>%
  summarise(
    `No F508del` = mean(genotype == "No F508del") * 100,
    `F508del heterozygous` = mean(genotype == "F508del heterozygous") * 100,
    `F508del homozygous` = mean(genotype == "F508del homozygous") * 100
  ) %>%
  pivot_longer(c(
    `No F508del`, `F508del heterozygous`, `F508del homozygous`
  ), names_to = "genotype", values_to = "perc") %>%
  ggplot(aes(x = factor(admit_year), y = perc, fill = factor(genotype))) +
  geom_bar(position = "stack", stat = "identity") +
  geom_text(aes(label = paste(round(perc, 1), "%")),
    position = position_stack(vjust = 0.5), size = 2
  ) +
  ylab("%") +
  facet_grid(~redcap_data_access_group) +
  theme_bw() +
  theme(legend.title = element_blank(), axis.title.x = element_blank()) +
  scale_fill_manual(values = c(red, purple, yellow))
```

## Slide 12

For people with multiple admissions, FEV1 % predicted was averaged at admission and at discharge. Baseline FEV1 % predicted is the average of up to 6 months of FEV1 measures prior to first admission.

```{r}
#| warning: false
df %>% group_by(redcap_data_access_group,admit_year,sid) %>%
  summarise(across(ppfev1_admit:ppfev1_baseline,~mean(.x,na.rm=T)),
            .groups = "drop") %>%
  rename(Admission = ppfev1_admit,Discharge = ppfev1_discharge,
         Baseline = ppfev1_baseline) %>%
  pivot_longer(c(Admission, Discharge, Baseline)) %>%
  mutate(name = factor(name,levels=c("Baseline","Admission", "Discharge"))) %>%
  ggplot(aes(x = name, y = value, 
             fill = factor(admit_year))) +
  geom_boxplot() + ylab("FEV1 % predicted") +
  facet_wrap(~redcap_data_access_group)+
  theme_bw() +
  theme(legend.title = element_blank(),legend.position = "bottom",
        axis.title.x = element_blank()) +
  scale_fill_manual(values = c(purple, yellow))
```
